---
title: "Exercise 2 - Random Number Generation through CDF and acceptance-rejection sampling"
author: "Florescu Andrei - 12242987"
date: "2024"
output: pdf_document
---

## Table of contents

- [Statistical Simulation project index](../README.md)

# 1. Linear Congruential Random Number Generation Algorithm

```{r, echo=TRUE}
mc.gen <- function(n, m, a, c = 0, x0) {
    us <- numeric(n)
    for (i in 1:n) {
        x0 <- a * (x0 + c) %% m
        us[i] <- x0 / m
    }
    return(us)
}

seed <- 12242987

round(mc.gen(8, 17, 3, 0, seed), 4)
round(mc.gen(15, 11, 3, 0, seed), 4)
round(mc.gen(15, 12451, 110, 0, seed), 4)
round(mc.gen(15, 9, 3, 0, seed), 4)
round(mc.gen(15, 10, 3, 0, seed), 4)
```

We can see that for certain values of m and a the randomly generated samples either repeat,
or either they have very predictable values (i.e. with one decmial).

# 2. Exponential Distribution

## Mathematical formula

Using the inverse transform sampling method we can solve for X in the exponential 
Distribution formula. Therefore the formula becomes: 

$$
F(x) = 1 - \exp(-\lambda x)
$$
We want the generated number to be uniform (between 0 and 1)
$$
U = F(X) = 1 - \exp(-\lambda X)
$$

$$
\exp(-\lambda X) = 1 - U
$$
Apply natural logarithm on both sides.
$$
X = -\frac{1}{\lambda} \ln(1 - U)
$$

Since U is between 0 and 1 we can rewrite the formula without 1 - U since its the same

$$
X = -\frac{1}{\lambda} \ln(U)
$$

## R implementation

```{r, echo=TRUE}
exp.rng <- function(n, lambda) {
    X <- numeric(n)
    uniform_vars <- runif(n)
    for (i in 1:n) {
        X[i] <- -1.0 / lambda * log(uniform_vars[i])
    }
    return(X)
}

samples1 <- exp.rng(1000, 1.2345)
samples2 <- exp.rng(1000, 12.345)
samples3 <- exp.rng(1000, 33.333)
real_data1 <- rexp(1000, 1.2345)
real_data2 <- rexp(1000, 12.345)
real_data3 <- rexp(1000, 33.333)

qqplot(real_data1, samples1, main = "QQ-Plot for lambda = 0.12345")
abline(0, 1, col = "red")

qqplot(real_data2, samples2, main = "QQ-Plot for lambda = 1234.5")
abline(0, 1, col = "red")

qqplot(real_data3, samples3, main = "QQ-Plot for lambda = 33.333")
abline(0, 1, col = "red")
```

# 3. Beta Distribution

## Beta function where alpha and beta are 2

For the beta distribution where alpha and beta are 2, the function is really close
to a uniform distribution.

```{r, echo=TRUE}
curve(dbeta(x, 2, 2), from = 0, to = 1)
```

We can see that the maximum value is for x = 0.5

```{r, echo=TRUE}
dbeta(0.5, 2, 2)
```

Therefore 1.5 is a good constant to use with the uniform distribution for alpha and beta equal to 2


```{r, echo=TRUE}
beta.rng22 <- function(n) {
    x <- numeric(n)
    accepted <- 0
    iter <- 0
    c <- 1.5
    while (accepted < n) {
        u <- runif(1)
        iter <- iter + 1
        y <- runif(1)
        f <- dbeta(y, 2, 2)

        if (u <= f / (c * y)) {
            accepted <- accepted + 1
            x[accepted] <- y
        }
    }
    return(x)
}

samples <- beta.rng22(1000)
real_data <- rbeta(1000, 2, 2)
qqplot(real_data, samples, main = "QQ-Plot for Beta with alpha=beta=2")
abline(0, 1, col = "red")
```

## Generalization

For a function that can have alpha and beta greater than one, we can still sample from the unfiorm distribution
The maximum value is always going to be at x=0.5 therefore we can calculate the constant
at the beginning of the function as c=beta(0.5,alpha,beta). 

```{r, echo=TRUE}
beta.rng <- function(n, alpha, beta) {
    x <- numeric(n)
    accepted <- 0
    iter <- 0
    c <- dbeta(0.5, alpha, beta)
    while (accepted < n) {
        u <- runif(1)
        iter <- iter + 1
        y <- runif(1)
        f <- dbeta(y, alpha, beta)

        if (u <= f / (c * y)) {
            accepted <- accepted + 1
            x[accepted] <- y
        }
    }
    return(x)
}

samples <- beta.rng(1000, 10, 10)
real_data <- rbeta(1000, 10, 10)
qqplot(real_data, samples, main = "QQ-Plot for Beta with alpha=beta=10")
abline(0, 1, col = "red")

samples <- beta.rng(1000, 100, 100)
real_data <- rbeta(1000, 100, 100)
qqplot(real_data, samples, main = "QQ-Plot for Beta with alpha=beta=100")
abline(0, 1, col = "red")
```

One could argue that we could use the normal distribution when alpha and beta are very large because in the 
Relations between Distributions visualised slide in the course, for alpha and beta closing in on infinity,
the beta distribution is closer to the normal distribution.


