---
title: "Exercise 6 - Cross Validation of Models"
author: "Florescu Andrei - 12242987"
date: "2024"
output:
  pdf_document:
    latex_engine: xelatex
---

## Table of contents

- [Statistical Simulation project index](../README.md)

# Task 1
```{r, echo=TRUE}
library(ISLR)
library(knitr)

data(Auto)
set.seed(12242987)

kable(head(Auto), caption = "Auto dataset (head)")
```

## 1. Fit multiple models
```{r, echo=TRUE}
# fit models
model1 <- lm(mpg ~ horsepower, data = Auto)
model2 <- lm(mpg ~ poly(horsepower, 2), data = Auto)
model3 <- lm(mpg ~ poly(horsepower, 3), data = Auto)

# predict
hp_seq <- seq(min(Auto$horsepower), max(Auto$horsepower), length.out = 500)
pred1 <- predict(model1, newdata = data.frame(horsepower = hp_seq))
pred2 <- predict(model2, newdata = data.frame(horsepower = hp_seq))
pred3 <- predict(model3, newdata = data.frame(horsepower = hp_seq))
model_preds <- data.frame(
  horsepower = hp_seq,
  model1 = pred1,
  model2 = pred2,
  model3 = pred3
)

library(ggplot2)
library(tidyr)
model_preds_long <- pivot_longer(model_preds, cols = -horsepower, names_to = "Model", values_to = "mpg")

# plot
ggplot(Auto, aes(x = horsepower, y = mpg)) +
  geom_point(color = "blue", alpha = 0.6) + # scatterplot
  geom_line(data = model_preds_long, aes(x = horsepower, y = mpg, color = Model), linewidth = 1) + # model lines
  labs(
    title = "Models Comparison on MPG vs Horsepower",
    x = "horsepower",
    y = "mpg",
    color = "Model"
  ) +
  theme_minimal()
```

## 2. Use the validation set approach to compare the models for 50/50 split and 70/30 split.
```{r, echo=TRUE}
compute_metrics <- function(actual, predicted) {
  mse <- mean((actual - predicted)^2)
  rmse <- sqrt(mse)
  mad <- median(abs(actual - predicted))
  list(MSE = mse, RMSE = rmse, MAD = mad)
}
```
### Split the data using a 50/50 split
```{r, echo=TRUE}
# split data
n <- nrow(Auto)
split_ratio <- 0.5
train_indices <- sample(1:n, size = floor(split_ratio * n))
train_data <- Auto[train_indices, ]
test_data <- Auto[-train_indices, ]

# train
model1 <- lm(mpg ~ horsepower, data = train_data)
model2 <- lm(mpg ~ poly(horsepower, 2), data = train_data)
model3 <- lm(mpg ~ poly(horsepower, 3), data = train_data)
pred1 <- predict(model1, newdata = test_data)
pred2 <- predict(model2, newdata = test_data)
pred3 <- predict(model3, newdata = test_data)

# comute metrics
metrics1 <- compute_metrics(test_data$mpg, pred1)
metrics2 <- compute_metrics(test_data$mpg, pred2)
metrics3 <- compute_metrics(test_data$mpg, pred3)

metrics_df_50 <- data.frame(
  Model = c("Linear", "Quadratic", "Cubic"),
  MSE = c(metrics1$MSE, metrics2$MSE, metrics3$MSE),
  RMSE = c(metrics1$RMSE, metrics2$RMSE, metrics3$RMSE),
  MAD = c(metrics1$MAD, metrics2$MAD, metrics3$MAD)
)
kable(metrics_df_50, caption = "Model metrics comparison 50/50 split")
```

For the 50/50 split we can see based on the mse, rmse and mad that the best model is 
the cubic model because it has the lowest metrics

### Split the data using a 70/30 split
```{r, echo=TRUE}
# split data
n <- nrow(Auto)
split_ratio <- 0.7
train_indices <- sample(1:n, size = floor(split_ratio * n))
train_data <- Auto[train_indices, ]
test_data <- Auto[-train_indices, ]

# train
model1 <- lm(mpg ~ horsepower, data = train_data)
model2 <- lm(mpg ~ poly(horsepower, 2), data = train_data)
model3 <- lm(mpg ~ poly(horsepower, 3), data = train_data)
pred1 <- predict(model1, newdata = test_data)
pred2 <- predict(model2, newdata = test_data)
pred3 <- predict(model3, newdata = test_data)

# comute metrics
metrics1 <- compute_metrics(test_data$mpg, pred1)
metrics2 <- compute_metrics(test_data$mpg, pred2)
metrics3 <- compute_metrics(test_data$mpg, pred3)

metrics_df_70 <- data.frame(
  Model = c("Linear", "Quadratic", "Cubic"),
  MSE = c(metrics1$MSE, metrics2$MSE, metrics3$MSE),
  RMSE = c(metrics1$RMSE, metrics2$RMSE, metrics3$RMSE),
  MAD = c(metrics1$MAD, metrics2$MAD, metrics3$MAD)
)
kable(metrics_df_70, caption = "Model metrics comparison 70/30 split")
```

For the 70/30 split the best model is the Quadratic model because it has the lowest 
score on two out of the three metrics, and is followed closely in performance 
by the cubic model.

## 3. Use the cv.glm function in the boot package for leave one out cross validation and k fold cross validation
```{r, echo=TRUE}
library(boot)
model1 <- glm(mpg ~ horsepower, data = Auto)
model2 <- glm(mpg ~ poly(horsepower, 2), data = Auto)
model3 <- glm(mpg ~ poly(horsepower, 3), data = Auto)
```

Leave one out cross Validation

```{r, echo=TRUE}
loocv_model1 <- cv.glm(Auto, model1)
loocv_model2 <- cv.glm(Auto, model2)
loocv_model3 <- cv.glm(Auto, model3)

loocv_results <- data.frame(
  Model = c("Linear", "Quadratic", "Cubic"),
  Error = c(loocv_model1$delta[1], loocv_model2$delta[1], loocv_model3$delta[1])
)
kable(loocv_results, caption = "Leave one out cross validation metrics comparison")
```

In this scenario we can conclude that the best model would be the quadratic model

### 5-fold and 10-fold cross validation

```{r, echo=TRUE}
kfold5_model1 <- cv.glm(Auto, model1, K = 5)
kfold5_model2 <- cv.glm(Auto, model2, K = 5)
kfold5_model3 <- cv.glm(Auto, model3, K = 5)

kfold10_model1 <- cv.glm(Auto, model1, K = 10)
kfold10_model2 <- cv.glm(Auto, model2, K = 10)
kfold10_model3 <- cv.glm(Auto, model3, K = 10)

kfold_results <- data.frame(
  Model = c("Linear", "Quadratic", "Cubic"),
  `5-Fold CV Error` = c(kfold5_model1$delta[1], kfold5_model2$delta[1], kfold5_model3$delta[1]),
  `10-Fold CV Error` = c(kfold10_model1$delta[1], kfold10_model2$delta[1], kfold10_model3$delta[1])
)
kable(kfold_results, caption = "K fold cross validation metrics comparison")
```

In the case of 5-fold CV we can say that the best model is the quadratic model, but for
10-fold validation the best model would be the cubic model

## 4. Compare all results from 2 and 3. in a table and draw your conclusions.

```{r, echo=TRUE}
kable(metrics_df_50, caption = "Model metrics comparison 50/50 split")
kable(metrics_df_70, caption = "Model metrics comparison 70/30 split")
kable(loocv_results, caption = "Leave one out cross validation metrics comparison")
kable(kfold_results, caption = "K fold cross validation metrics comparison")
```

Overall in most cases the best model is the Quadratic model, followed very closely in 
performance by the cubic model. The worst performing model was the linear model

# Task 2
```{r, echo=TRUE}
data(economics)
kable(head(economics), caption = "economics dataset (head)")
```

## 1. Fit multiple models

### Linear models
```{r, echo=TRUE}
lm1 <- lm(unemploy ~ uempmed, data = economics)
summary(lm1)
lm2 <- lm(uempmed ~ unemploy, data = economics)
summary(lm2)
```

### Exponential and logarithmic models
```{r, echo=TRUE}
log_lm1 <- lm(log(unemploy) ~ uempmed, data = economics)
summary(log_lm1)

log_lm2 <- lm(log(uempmed) ~ unemploy, data = economics)
summary(log_lm2)

exp_lm1 <- lm(uempmed ~ log(unemploy), data = economics)
summary(exp_lm1)

exp_lm2 <- lm(unemploy ~ log(uempmed), data = economics)
summary(exp_lm2)
```

### Polynomial models

```{r, echo=TRUE}
# models for unemploy ~ uempmed
poly2_lm1 <- lm(unemploy ~ poly(uempmed, 2), data = economics)
summary(poly2_lm1)
poly3_lm1 <- lm(unemploy ~ poly(uempmed, 3), data = economics)
summary(poly3_lm1)
poly10_lm1 <- lm(unemploy ~ poly(uempmed, 10), data = economics)
summary(poly10_lm1)

# models for uempmed ~ unemploy
poly2_lm2 <- lm(uempmed ~ poly(unemploy, 2), data = economics)
summary(poly2_lm2)

poly3_lm2 <- lm(uempmed ~ poly(unemploy, 3), data = economics)
summary(poly3_lm2)

poly10_lm2 <- lm(uempmed ~ poly(unemploy, 10), data = economics)
summary(poly10_lm2)
```

## 2. Plot the models
### Plot for unemploy ~ uempmed models
```{r, echo=TRUE}
library(tidyr)

uempmed_seq <- seq(min(economics$uempmed), max(economics$uempmed), length.out = 500)
df <- data.frame(uempmed = uempmed_seq)

model_preds <- data.frame(
  uempmed = uempmed_seq,
  lm1 = predict(lm1, newdata = df),
  log_lm1 = predict(log_lm1, newdata = df),
  exp_lm2 = predict(exp_lm2, newdata = df),
  poly2_lm1 = predict(poly2_lm1, newdata = df),
  poly3_lm1 = predict(poly3_lm1, newdata = df),
  poly10_lm1 = predict(poly10_lm1, newdata = df)
)

model_preds_long <- pivot_longer(model_preds, cols = -uempmed, names_to = "Model", values_to = "unemploy")

# Plot
ggplot(economics, aes(x = uempmed, y = unemploy)) +
  geom_point(color = "blue", alpha = 0.6) + # Scatter plot
  geom_line(data = model_preds_long, aes(x = uempmed, y = unemploy, color = Model), linewidth = 1) + # Model lines
  labs(
    title = "Models Comparison on unemploy ~ uempmed",
    x = "uempmed",
    y = "unemploy",
    color = "Model"
  ) +
  theme_minimal()
```

Based on the graph i think the best model could be the exponential model

### Plot for uempmed ~ unemploy models

```{r, echo=TRUE}
library(tidyr)

unemploy_seq <- seq(min(economics$unemploy), max(economics$unemploy), length.out = 500)
df <- data.frame(unemploy = unemploy_seq)

model_preds <- data.frame(
  unemploy = unemploy_seq,
  lm2 = predict(lm2, newdata = df),
  log_lm2 = predict(log_lm2, newdata = df),
  exp_lm1 = predict(exp_lm1, newdata = df),
  poly2_lm2 = predict(poly2_lm2, newdata = df),
  poly3_lm2 = predict(poly3_lm2, newdata = df),
  poly10_lm2 = predict(poly10_lm2, newdata = df)
)

model_preds_long <- pivot_longer(model_preds, cols = -unemploy, names_to = "Model", values_to = "uempmed")

# Plot
ggplot(economics, aes(x = unemploy, y = uempmed)) +
  geom_point(color = "blue", alpha = 0.6) + # Scatter plot
  geom_line(data = model_preds_long, aes(x = unemploy, y = uempmed, color = Model), linewidth = 1) + # Model lines
  labs(
    title = "Models Comparison on uempmed ~ unemploy",
    x = "unemploy",
    y = "uempmed",
    color = "Model"
  ) +
  theme_minimal()
```

Based on the graph i think the best model could be the cubic polynomial model

## 3. Use the cv.glm function in the boot package
```{r, echo=TRUE}
cv_rmse_mse <- function(model, data, K) {
  cv_result <- cv.glm(data, model, K = K)
  mse <- mean(cv_result$delta)
  rmse <- sqrt(mse)
  return(list(MSE = mse, RMSE = rmse))
}

models <- list(
  lm1 = glm(unemploy ~ uempmed, data = economics),
  lm2 = glm(uempmed ~ unemploy, data = economics),
  log_lm1 = glm(log(unemploy) ~ uempmed, data = economics),
  log_lm2 = glm(log(uempmed) ~ unemploy, data = economics),
  exp_lm1 = glm(uempmed ~ log(unemploy), data = economics),
  exp_lm2 = glm(unemploy ~ log(uempmed), data = economics),
  poly2_lm1 = glm(unemploy ~ poly(uempmed, 2), data = economics),
  poly3_lm1 = glm(unemploy ~ poly(uempmed, 3), data = economics),
  poly10_lm1 = glm(unemploy ~ poly(uempmed, 10), data = economics),
  poly2_lm2 = glm(uempmed ~ poly(unemploy, 2), data = economics),
  poly3_lm2 = glm(uempmed ~ poly(unemploy, 3), data = economics),
  poly10_lm2 = glm(uempmed ~ poly(unemploy, 10), data = economics)
)


```

### LOOCV results

```{r, echo=TRUE}
library(dplyr)
cv_results_loo <- lapply(models, function(model) {
  cv_rmse_mse(model, economics, K = nrow(economics))
})

cv_results_df <- bind_rows(
  lapply(names(cv_results_loo), function(model_name) {
    result <- cv_results_loo[[model_name]]
    data.frame(
      Model = model_name,
      MSE = result$MSE,
      RMSE = result$RMSE
    )
  })
)
kable(cv_results_df, caption = "LOOCV results")
```

We get the best performance with the logarithmic models

### 5 fold CV
```{r, echo=TRUE}
cv_results_5fold <- lapply(models, function(model) {
  cv_result <- cv.glm(economics, model, K = 5)  # 5-fold cross-validation
  mse <- mean(cv_result$delta)
  rmse <- sqrt(mse)
  return(list(MSE = mse, RMSE = rmse))
})

cv_results_df <- bind_rows(
  lapply(names(cv_results_5fold), function(model_name) {
    result <- cv_results_5fold[[model_name]]
    data.frame(
      Model = model_name,
      MSE = result$MSE,
      RMSE = result$RMSE
    )
  })
)
kable(cv_results_df, caption = "5 fold CV results")
```

We get the best performance with the logarithmic models

### 10 fold CV

```{r, echo=TRUE}
cv_results_10fold <- lapply(models, function(model) {
  cv_result <- cv.glm(economics, model, K = 10)  # 10-fold cross-validation
  mse <- mean(cv_result$delta)
  rmse <- sqrt(mse)
  return(list(MSE = mse, RMSE = rmse))
})

cv_results_df <- bind_rows(
  lapply(names(cv_results_10fold), function(model_name) {
    result <- cv_results_10fold[[model_name]]
    data.frame(
      Model = model_name,
      MSE = result$MSE,
      RMSE = result$RMSE
    )
  })
)
kable(cv_results_df, caption = "5 fold CV results")
```

We get the best performance with the logarithmic models and also some good performance 
with the polynomial models

## 4. Explain based on the CV and graphical model fits the concepts of Underfitting, Overfitting and how to apply cross-validation to determine the appropriate model fit.

Underfitting occurs when a model is too simple meaning it has not learned enough 
from the data or is too general, resulting in poor performance on both training 
and test data.

Overfitting occurs when a model is too complex meaning it has learned the training 
data too well and cannot generalize on unseen cases (test data), in other words 
performing with a very high accuracy on the training set but with a poor accuracy
on the testing set.

Cross validation allows us to test the models on different subsets of the data 
and is a good way to prevent Underfitting and Overfitting. LOOCV leaves only one entry
as testing data, while K fold CV leaves 1 fold out of the K folds as testing data and 
trains on the remaining K-1 folds. These methods were also very usefull in figuring 
out way faster which model fits the data better by having the ability to compute
the metrics we wanted very fast.
