---
title: "Exercise 3 - Monte Carlo Simulation of areas"
author: "Florescu Andrei - 12242987"
date: "2024"
output:
  pdf_document:
    latex_engine: xelatex
---

# Setup

```{r, echo=TRUE}
set.seed(12242987)
```

# Task 1

## 1. Approximate the integral for b = 6 (using Monte Carlo integration).

```{r, echo=TRUE}
a <- 1
b <- 6
n <- 100000
random_vars <- runif(n, min = a, max = b)

function_to_integrate <- function(x) {
    return(exp(-x^2))
}
```

Monte Carlo aproximation

```{r, echo=TRUE}
monte_carlo_aprox <- (b - a) * mean(function_to_integrate(random_vars))
print(monte_carlo_aprox)
```

Result using integrate function

```{r, echo=TRUE}
integrate_result <- integrate(function_to_integrate, lower = a, upper = b)$value
print(integrate_result)
```

## 2. Compute the integral for b = $\infty$

We can rewrite the integral  

\[
\int_1^{\infty} e^{-x^2} \, dx = \int_0^{\infty} e^{-(x+1)^2} \, dx
\]


then use the exponential distribution to aproximate the integral

```{r, echo=TRUE}
X <- rexp(n)
monte_carlo_aprox <- mean(exp(-(X + 1)^2) / dexp(X))
print(monte_carlo_aprox)
```

Aproximation using integrate function

```{r, echo=TRUE}
integrate(function_to_integrate, 1, Inf)
```

## 3. Do you have an explanation why Monte Carlo integration agrees in 2. with integrate but not so much in 1.?

The integral in the first example is close to the integrate function to the second decmial
and the integral in the second example is close to the third decimal, which is not a huge difference
but the reason might be due to the Strong Law of Large Numbers


# Task 2



## 1. r(t) function plot
```{r, echo=TRUE}
library(ggplot2)

r_func <- function(t) {
    return(exp(cos(t)) - 2 * cos(4 * t) - (sin(t / 12))^5)
}
t_values <- seq(-pi, pi, length.out = 10000)

polar_coord <- data.frame(
    t = t_values,
    x = r_func(t_values) * cos(t_values),
    y = r_func(t_values) * sin(t_values)
)

ggplot(polar_coord, aes(x = x, y = y)) +
    geom_polygon(fill = "skyblue", color = "blue", alpha = 0.5) +
    coord_fixed() +
    theme_minimal() +
    labs(
        title = "r(t)",
        x = "x",
        y = "y"
    ) +
    theme(plot.title = element_text(hjust = 0.5))
```

## 2. Generate random coordinates within the rectangle [−2,3.5]x[−3,3]


```{r, echo=TRUE}
n <- 1000 # Number of points
x <- runif(n, min = -2, max = 3.5)
y <- runif(n, min = -3, max = 3)

library(dplyr)

points_data <- data.frame(
    x = x,
    y = y
)

points_data <- points_data %>% mutate(
    t = atan2(y, x), # calculate t
    r_point = sqrt(x^2 + y^2), # calculate radius for each point
    r_boundary = r_func(t), # calculate boundary radius at angle t
    is_in_area = r_point <= r_boundary # check if point is in func area
)
head(points_data)
```


## 3. Simulate 100, 1000, 10000 and 100000 random coordiantes and calculate the percentage of the points within the enclosed area.

```{r, echo=TRUE}
get_points_percentages <- function(n) {
    x <- runif(n, min = -2, max = 3.5) # get n points
    y <- runif(n, min = -3, max = 3)
    points_data <- data.frame(
        x = x,
        y = y
    )
    # determine if points are in function area
    points_data <- points_data %>% mutate(
        t = atan2(y, x), # calculate t
        r_point = sqrt(x^2 + y^2), # calculate radius for each point
        r_boundary = r_func(t), # calculate boundary radius at angle t
        is_in_area = r_point <= r_boundary # check if point is in func area
    )
    percentage_inside <- table(points_data$is_in_area)
    return(
        percentage_inside["TRUE"] /
            (percentage_inside["TRUE"] + percentage_inside["FALSE"])
    )
}

get_points_percentages(n = 100)
get_points_percentages(n = 1000)
get_points_percentages(n = 10000)
get_points_percentages(n = 100000)
```

We can assume that the area is about 0.38 of the rectangle area defined by the limits
[−2,3.5]x[−3,3] which is:

```{r, echo=TRUE}
(3.5 + 2) * (3 + 3) * 0.38
```


```{r, echo=TRUE}
t <- seq(-pi, pi, length.out = 1000)
curve_data <- data.frame(
    x = r_func(t) * cos(t),
    y = r_func(t) * sin(t)
)
results <- lapply(c(100, 1000, 10000, 100000), get_points_percentages)
results_df <- do.call(rbind, lapply(results, as.data.frame))
rownames(results_df) <- c("100", "1000", "10000", "100000")
results_df["Area"] <- results_df["X[[i]]"] * (3.5 + 2) * (3 + 3)
colnames(results_df) <- c("Percentage", "Area")
results_df
```

## 4. Explain the functionality of Monte Carlo simulation in your own words referring to these simulations

The Monte Carlo simulation is a great tool that can be used to aproximate calculations 
in scenarios where direct computation might be difficult to perform(for example the integral mentioned in task 1)
It is also very useful in aproximating areas of complex functions by defining a simpler 
region that contains the entire area of the function and then estimate the area of the desired function by
checking the percentage of points inside the shape. The simulation result converges towards
the real value when the number of generations is large enough and therefore it is much reliable.

