---
title: "Exercise 5 - Sampling Intervals for Models"
author: "Florescu Andrei - 12242987"
date: "2024"
output:
  pdf_document:
    latex_engine: xelatex
---

# Task 1

```{r, echo=TRUE}
set.seed(12242987)
x1 <- c(-0.673, -0.584, 0.572, -0.341, -0.218, 0.603, -0.415, -0.013, 0.763, 0.804, 0.054, 1.746, -0.472, 1.638, -0.578, 0.947, -0.329, -0.188, 0.794, 0.894, -1.227, 1.059) # nolint
x2 <- c(0.913, -0.639, 2.99, -5.004, 3.118, 0.1, 1.128, 0.579, 0.32, -0.488, -0.994, -0.212, 0.413, 1.401, 0.007, 0.568, -0.005, 0.696) # nolint
```

## 1. Plot the data in a way which visualises the comparison of means appropriately.
```{r, echo=TRUE}
mean_x1 <- mean(x1)
mean_x2 <- mean(x2)

boxplot(list(x1 = x1, x2 = x2),
    col = c("lightblue", "lightgreen"),
    names = c("Sample 1 (x1)", "Sample 2 (x2)"), main = "Comparison of Means",
    ylab = "Values"
)
points(rep(1, length(x1)), x1, col = "blue", pch = 19, cex = 0.7)
points(rep(2, length(x2)), x2, col = "red", pch = 19, cex = 0.7)
abline(h = mean_x1, col = "blue", lty = 2)
abline(h = mean_x2, col = "red", lty = 2)
legend("topright",
    legend = c(
        paste("Mean x1:", round(mean_x1, 2)),
        paste("Mean x2:", round(mean_x2, 2))
    ),
    col = c("blue", "red"), lty = 2, cex = 0.8
)
```

## 2. Consider different sampling schemes Argue for choice what is more natural and which advantages or disadvantages may apply.
If we want to compare each group (which is specified in the hypthesis H1), then i think sampling 
with replacement from each group is more natural because this way we can see the differences between the groups
better. If the two groups are considered to be very similar or we assume that they are aproximately the same,
then I think the other method (Centering both samples and then resample from the 
combined samples x1 and x2 for n1 and n2 times) is more natural (if we consider proving H0).

## 3. Bootstrap using both strategies mentioned above using the t-test statistic. Calculate the bootstrap p-value based on 10000 bootstrap samples and 0.95 as well as 0.99 confidence intervals. Make your decision at the significance level 0.05 or 0.01, respectively.
```{r, echo=TRUE}
bootstrap_t_test_independent <- function(x1, x2, num_bootstrap = 10000, alpha = 0.05) {
    observed_t_stat <- t.test(x1, x2, var.equal = FALSE)$statistic
    n1 <- length(x1)
    n2 <- length(x2)

    bootstrap_t_stats <- numeric(num_bootstrap)

    for (i in 1:num_bootstrap) {
        resampled_x1 <- sample(x1, size = n1, replace = TRUE)
        resampled_x2 <- sample(x2, size = n2, replace = TRUE)
        bootstrap_t_stats[i] <- t.test(resampled_x1, resampled_x2, var.equal = FALSE)$statistic
    }

    p_value <- mean(abs(bootstrap_t_stats) >= abs(observed_t_stat))

    ci_lower <- quantile(bootstrap_t_stats, alpha / 2)
    ci_upper <- quantile(bootstrap_t_stats, 1 - alpha / 2)

    list(
        observed_t_stat = observed_t_stat,
        p_value = p_value,
        confidence_interval = c(ci_lower, ci_upper)
    )
}

bootstrap_t_test_pooled <- function(x1, x2, num_bootstrap = 10000, alpha = 0.05) {
    observed_t_stat <- t.test(x1, x2, var.equal = FALSE)$statistic
    n1 <- length(x1)
    n2 <- length(x2)

    x1_centered <- x1 - mean(x1)
    x2_centered <- x2 - mean(x2)
    pooled_data <- c(x1_centered, x2_centered)

    bootstrap_t_stats <- numeric(num_bootstrap)

    for (i in 1:num_bootstrap) {
        resampled_data <- sample(pooled_data, size = n1 + n2, replace = TRUE)
        resampled_x1 <- resampled_data[1:n1]
        resampled_x2 <- resampled_data[(n1 + 1):(n1 + n2)]
        bootstrap_t_stats[i] <- t.test(resampled_x1, resampled_x2, var.equal = FALSE)$statistic
    }

    p_value <- mean(abs(bootstrap_t_stats) >= abs(observed_t_stat))

    ci_lower <- quantile(bootstrap_t_stats, alpha / 2)
    ci_upper <- quantile(bootstrap_t_stats, 1 - alpha / 2)

    list(
        observed_t_stat = observed_t_stat,
        p_value = p_value,
        confidence_interval = c(ci_lower, ci_upper)
    )
}
results_independent <- bootstrap_t_test_independent(x1, x2, alpha = 0.05)
results_pooled <- bootstrap_t_test_pooled(x1, x2, alpha = 0.05)

results_independent_99 <- bootstrap_t_test_independent(x1, x2, alpha = 0.01)
results_pooled_99 <- bootstrap_t_test_pooled(x1, x2, alpha = 0.01)
```
```{r, echo=TRUE}
results_independent
```
```{r, echo=TRUE}
results_pooled
```
```{r, echo=TRUE}
results_independent_99
```
```{r, echo=TRUE}
results_pooled_99
```

Based we can reject H1 and say that there is no statistical significance between the 
two groups which means the results are consistent with the null hypothesis H0

## 4. Implement the corresponding permutation test and obtain p-value and confidence intervals as in 3. to get a corresponding test decision at the same significance levels.
```{r, echo=TRUE}
permutation_t_test <- function(x1, x2, num_permutations = 10000, alpha = 0.05) {
    observed_t_stat <- t.test(x1, x2, var.equal = FALSE)$statistic
    combined_data <- c(x1, x2)
    n1 <- length(x1)
    n2 <- length(x2)

    perm_t_stats <- numeric(num_permutations)

    for (i in 1:num_permutations) {
        permuted_data <- sample(combined_data)
        perm_x1 <- permuted_data[1:n1]
        perm_x2 <- permuted_data[(n1 + 1):(n1 + n2)]
        perm_t_stats[i] <- t.test(perm_x1, perm_x2, var.equal = FALSE)$statistic
    }

    p_value <- mean(abs(perm_t_stats) >= abs(observed_t_stat))
    ci_lower <- quantile(perm_t_stats, alpha / 2)
    ci_upper <- quantile(perm_t_stats, 1 - alpha / 2)

    list(
        observed_t_stat = observed_t_stat,
        p_value = p_value,
        confidence_interval = c(ci_lower, ci_upper)
    )
}
```

```{r, echo=TRUE}
results_permutation <- permutation_t_test(x1, x2, alpha = 0.05)
results_permutation_99 <- permutation_t_test(x1, x2, alpha = 0.01)
results_permutation
results_permutation_99
```
We can see that the result doesn\'t change, we still reject H1

## 5. The Wilcoxon rank sum test statistic is the sum of ranks of the observations of sample 1 computed in the combined sample. Use bootstrapping with both strategies mentioned above and obtain p-value and confidence intervals as in 3. to get a corresponding test decision at the same significance levels.
```{r, echo=TRUE}
bootstrap_wilcoxon_independent <- function(x1, x2, num_bootstrap = 10000, alpha = 0.05) {
    combined_data <- c(x1, x2)
    ranks <- rank(combined_data)
    observed_stat <- sum(ranks[1:length(x1)])

    n1 <- length(x1)
    n2 <- length(x2)
    bootstrap_stats <- numeric(num_bootstrap)

    for (i in 1:num_bootstrap) {
        resampled_x1 <- sample(x1, size = n1, replace = TRUE)
        resampled_x2 <- sample(x2, size = n2, replace = TRUE)
        combined_resample <- c(resampled_x1, resampled_x2)
        resample_ranks <- rank(combined_resample)
        bootstrap_stats[i] <- sum(resample_ranks[1:n1])
    }
    p_value <- mean(abs(bootstrap_stats - mean(bootstrap_stats)) >= abs(observed_stat - mean(bootstrap_stats)))
    ci_lower <- quantile(bootstrap_stats, alpha / 2)
    ci_upper <- quantile(bootstrap_stats, 1 - alpha / 2)

    list(
        observed_stat = observed_stat,
        p_value = p_value,
        confidence_interval = c(ci_lower, ci_upper)
    )
}

bootstrap_wilcoxon_pooled <- function(x1, x2, num_bootstrap = 10000, alpha = 0.05) {
    combined_data <- c(x1, x2)
    ranks <- rank(combined_data)
    observed_stat <- sum(ranks[1:length(x1)])

    n1 <- length(x1)
    n2 <- length(x2)
    bootstrap_stats <- numeric(num_bootstrap)

    pooled_data <- c(x1, x2)

    for (i in 1:num_bootstrap) {
        resampled_data <- sample(pooled_data, size = n1 + n2, replace = TRUE)
        resample_ranks <- rank(resampled_data)
        bootstrap_stats[i] <- sum(resample_ranks[1:n1])
    }
    p_value <- mean(abs(bootstrap_stats - mean(bootstrap_stats)) >= abs(observed_stat - mean(bootstrap_stats)))
    ci_lower <- quantile(bootstrap_stats, alpha / 2)
    ci_upper <- quantile(bootstrap_stats, 1 - alpha / 2)

    list(
        observed_stat = observed_stat,
        p_value = p_value,
        confidence_interval = c(ci_lower, ci_upper)
    )
}
```

```{r, echo=TRUE}
results_wilcoxon_independent <- bootstrap_wilcoxon_independent(x1, x2, alpha = 0.05)
results_wilcoxon_pooled <- bootstrap_wilcoxon_pooled(x1, x2, alpha = 0.05)

results_wilcoxon_independent_99 <- bootstrap_wilcoxon_independent(x1, x2, alpha = 0.01)
results_wilcoxon_pooled_99 <- bootstrap_wilcoxon_pooled(x1, x2, alpha = 0.01)

results_wilcoxon_independent
results_wilcoxon_pooled
results_wilcoxon_independent_99
results_wilcoxon_pooled_99
```

## 6. Compare results to the results using t.test and wilcox.test
```{r, echo=TRUE}
t_test_result <- t.test(x1, x2, var.equal = FALSE)
t_test_result

wilcoxon_test_result <- wilcox.test(x1, x2)
wilcoxon_test_result
```

We can see that the results of the tests are similar to our results for the wilcoxon test
but for the t test they are different.

# Task 2

## 1. Create a sample of size 200 from the model above and for the independent predictor x3
```{r, echo=TRUE}
library(knitr)

n <- 200
x1 <- rnorm(n, mean = 2, sd = sqrt(3))
x2 <- runif(n, min = 2, max = 4)
x3 <- runif(n, min = -2, max = 2)
epsilon <- rt(n, df = 5)

y <- 3 + 2 * x1 + x2 + epsilon
data <- data.frame(x1 = x1, x2 = x2, x3 = x3, y = y)
kable(head(data), caption = "Generated sample for model y")
```

## 2. Do residual bootstrap for linear regression and fit the model y∼x1+x2+x3. Get the percentile CI for the coefficients. Can you exclude x3?
Fit the linear regression model 
```{r, echo=TRUE}
model <- lm(y ~ x1 + x2 + x3, data = data)
```

Get the residuals
```{r, echo=TRUE}
residuals <- resid(model)
fitted_values <- fitted(model)
```

Boostraping
```{r, echo=TRUE}
B <- 1000 # number of bootstrap iterations

bootstrap_coefs <- matrix(NA, nrow = B, ncol = length(coef(model)))
colnames(bootstrap_coefs) <- names(coef(model))

for (i in 1:B) {
    resampled_residuals <- sample(residuals, replace = TRUE)
    new_y <- fitted_values + resampled_residuals
    new_model <- lm(new_y ~ x1 + x2 + x3, data = data)
    bootstrap_coefs[i, ] <- coef(new_model)
}
```

Calculate the confidence intervals and check if x3 can be excluded
```{r, echo=TRUE}
CI_lower <- apply(bootstrap_coefs, 2, function(x) quantile(x, probs = 0.025))
CI_upper <- apply(bootstrap_coefs, 2, function(x) quantile(x, probs = 0.975))

bootstrap_CI <- data.frame(
    Coefficient = names(coef(model)),
    Estimate = coef(model),
    CI_Lower = CI_lower,
    CI_Upper = CI_upper
)

kable(bootstrap_CI, caption = "Percentiles summary for residual boostrap")

exclude_x3 <- (CI_lower["x3"] <= 0 && CI_upper["x3"] >= 0)
exclude_x3
```

This means that x3 can be excluded from the model.

## 3. Do pairs bootstrap for linear regression and fit the model y∼x1+x2+x3. Get the percentile CI for the coefficients. Can you exclude x3?

We can use the previous base model we computed in 2. and the residuals from that model to do 
pairs boostraping
```{r, echo=TRUE}
n <- nrow(data)
for (i in 1:B) {
    bootstrap_indices <- sample(1:n, n, replace = TRUE)
    bootstrap_data <- data[bootstrap_indices, ]
    new_model <- lm(y ~ x1 + x2 + x3, data = bootstrap_data)
    bootstrap_coefs[i, ] <- coef(new_model)
}

CI_lower <- apply(bootstrap_coefs, 2, function(x) quantile(x, probs = 0.025))
CI_upper <- apply(bootstrap_coefs, 2, function(x) quantile(x, probs = 0.975))

bootstrap_CI <- data.frame(
    Coefficient = names(coef(model)),
    Estimate = coef(model),
    CI_Lower = CI_lower,
    CI_Upper = CI_upper
)

kable(bootstrap_CI, caption = "Percentiles summary for pairs boostrap")
exclude_x3 <- (CI_lower["x3"] <= 0 && CI_upper["x3"] >= 0)
exclude_x3
```

We can still exclude x3 from the model.

## 4. Compare the two approaches in 2. and 3. and explain the differences in the sampling approach and how this (might) affect(s) the results.

There are no major differences in the results between the two approaches, however,
using only the residuals this method might not be as accurate as the pairs bootstrap 
which resamples the entire dataset and therefore the confidence intervals and the 
result might be more sensitive than the first approach.

# Task 3

The boostraping methodology can be summarized as follows: 

1. Resample the data: Given a sample of size n, randomly select n data points
2. Fit the model: compute the statistic of interest or estimate the parameter of interest
3. Repeat the process a number of times (in our cases 10000 times)
4. Compute the confidence intervals and reject or accept the hypothesis

Boostraping is a very useful method because its easy to implement and it helps 
provide confidence intervals which are reliable and accurate. Some disadvantages are that the 
method can be computationaly heavy if there is a lot of data and a lot of samples to
compute, and since it uses only observed data it might be biased and sometimes not represent the actual
distribution.