---
title: "Exercise 7 - Comparing penalized regression estimators"
author: "Florescu Andrei - 12242987"
date: "2024"
output:
  pdf_document:
    latex_engine: xelatex
---

## Table of contents

- [Statistical Simulation project index](../README.md)

# Task 1
```{r, echo=TRUE}
set.seed(12242987)
library(knitr)
```

## 1. Write your own function for the lasso using the shooting algorithm
```{r, echo=TRUE}
soft_threshold <- function(z, gamma) {
  return(sign(z) * max(0, abs(z) - gamma))
}

lasso_shooting <- function(X, y, lambda, tol = 0.000001, max_iter = 1000) {
  n <- nrow(X)
  p <- ncol(X)
  X_mean <- colMeans(X)
  y_mean <- mean(y)
  X_centered <- scale(X, center = TRUE, scale = FALSE)
  y_centered <- y - y_mean

  # initialize coefs
  beta <- rep(0, p)
  beta_old <- beta

  #### Shooting algorithm ####

  # compute a_j (2 * sum(x_ij^2))
  a <- 2 * colSums(X_centered^2)

  for (iter in 1:max_iter) {
    for (j in 1:p) {
      # compute c_j
      r <- y_centered - X_centered %*% beta + beta[j] * X_centered[, j]
      c_j <- 2 * sum(X_centered[, j] * r)

      # soft update
      beta[j] <- soft_threshold(c_j / a[j], lambda / a[j])
    }

    # check for stopping
    if (max(abs(beta - beta_old)) < tol) {
      break
    }
    beta_old <- beta
  }

  # compute intercept
  intercept <- y_mean - sum(X_mean * beta)

  # return coefficients, intercept and nr of iterations
  return(list(intercept = intercept, coefficients = beta, iterations = iter))
}

## test to see if it works
n <- 100
p <- 5
X <- matrix(rnorm(n * p), n, p)
true_beta <- c(1, -5, -3, 2, 1) # random values selected as coefs
y <- X %*% true_beta + rnorm(n)
lambda <- 0.5
result <- lasso_shooting(X, y, lambda)
result
```

## 2. Write a function which computes the lasso using your algorithm for a vector of lambdas
```{r, echo=TRUE}
lasso_for_lamda_vector <- function(X, y, lambda_vector, tol = 0.000001, max_iter = 1000) {
  beta_matrix <- matrix(0, nrow = ncol(X)+1, ncol = length(lambda_vector))
  for (i in 1:length(lambda_vector)) { # nolint: seq_linter.
    lambda <- lambda_vector[i]
    result <- lasso_shooting(X, y, lambda, tol, max_iter)
    beta_matrix[, i] <- c(result$intercept, result$coefficients)
  }
  return(list(coefficients = beta_matrix, lambda = lambda_vector))
}

# test for a sample lambda vector
lambda_vector <- c(0.1, 0.3, 0.5, 0.7, 0.9)
result <- lasso_for_lamda_vector(X, y, lambda_vector)
result

```

## 3. Compare the performance and output of your functions against the lasso implementation from glmnet. 
```{r, echo=TRUE}
library(glmnet)
library(microbenchmark)
shooting_result <- lasso_for_lamda_vector(X, y, lambda_vector)

# Fit LASSO Using glmnet
glmnet_result <- glmnet(X, y, alpha = 1, lambda = lambda_vector, standardize = TRUE, intercept = TRUE)

# Compare Coefficients
cat("Comparison of Coefficients (Shooting vs glmnet):\n")
shooting_coef <- shooting_result$coefficients[-1, ] # Exclude intercept
glmnet_coef <- as.matrix(coef(glmnet_result))[-1, ] # Exclude intercept
```
### Shooting lasso coefficients
```{r, echo=TRUE}
shooting_coef
```
### glmnet coefficients
```{r, echo=TRUE}
glmnet_coef
```
### Execution time comparison
```{r, echo=TRUE}
benchmark_result <- microbenchmark(
  Shooting = lasso_for_lamda_vector(X, y, lambda_vector),
  glmnet = glmnet(X, y, alpha = 1, lambda = lambda_vector, standardize = TRUE, intercept = TRUE),
  times = 10
)
benchmark_result
```

## 4. Write a function to perform 10-fold cross-validation for the lasso using MSE as the performance measure. 

### lasso shooting results
```{r, echo=TRUE}
cv_lasso_shooting <- function(X, y, lambda_vec, folds = 10, tol = 1e-6, max_iter = 1000) {
  n <- nrow(X)
  fold_ids <- sample(rep(1:folds, length.out = n)) # random fold ids
  n_lambda <- length(lambda_vec)
  mse_matrix <- matrix(0, nrow = folds, ncol = n_lambda)

  ### CV ###
  for (k in 1:folds) {
    # split data
    train_index <- which(fold_ids != k)
    test_index <- which(fold_ids == k)

    X_train <- X[train_index, , drop = FALSE]
    y_train <- y[train_index]
    X_test <- X[test_index, , drop = FALSE]
    y_test <- y[test_index]

    # calculate lasso
    shooting_result <- lasso_for_lamda_vector(X_train, y_train, lambda_vec, tol, max_iter)

    # get coefs and intercept
    betas <- shooting_result$coefficients[-1, ] # Coefficients
    intercepts <- shooting_result$coefficients[1, ] # Intercepts

    # predict on test set and calculate mse
    for (l in 1:n_lambda) {
      y_pred <- intercepts[l] + X_test %*% betas[, l]
      mse_matrix[k, l] <- mean((y_test - y_pred)^2)
    }
  }

  # average mse at the end
  avg_mse <- colMeans(mse_matrix)

  # get lambda with lowest mse and rmse
  lambda_min_mse <- lambda_vec[which.min(avg_mse)]
  rmse <- sqrt(avg_mse)
  lambda_min_rmse <- lambda_vec[which.min(rmse)]

  # plot mse vs Lambda
  plot(log(lambda_vec), avg_mse,
    type = "b", pch = 19, col = "blue",
    xlab = "log(Lambda)", ylab = "Mean Squared Error (MSE)", main = "10-Fold Cross-Validation"
  )
  abline(v = log(lambda_min_mse), col = "red", lty = 2)
  abline(v = log(lambda_min_rmse), col = "green", lty = 2)
  legend("topright",
    legend = c("MSE", "Min MSE", "Min RMSE"),
    col = c("blue", "red", "green"), lty = c(1, 2, 2), pch = c(19, NA, NA), cex = 0.8
  )

  # return results
  list(
    lambda_min_mse = lambda_min_mse,
    lambda_min_rmse = lambda_min_rmse,
    mse = avg_mse,
    rmse = rmse,
    lambda = lambda_vec
  )
}

cv_result <- cv_lasso_shooting(X, y, lambda_vector)
cv_result

lambda_vec <- seq(0.1, 1, by = 0.1)
cv_result <- cv_lasso_shooting(X, y, lambda_vec)
cv_result
```

### glmnet results

```{r, echo=TRUE}
cv_glmnet <- cv.glmnet(X, y, alpha = 1, lambda = lambda_vec, nfolds = 10)
plot(cv_glmnet, main = "glmnet cv")
```

# Task 2
```{r, echo=TRUE}
library(ISLR)
# open dataset
data("Hitters")
df <- na.omit(Hitters)
y <- df$Salary
X <- model.matrix(Salary ~ ., data = df)[, -1]

# split data 70/30
train_index <- sample(1:n, size = 0.7 * n)
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

kable(head(df), caption = "Hitters dataset (head)")
```

## 1. Use your lasso function to decide which lambda is best here. Plot also the whole path for the coefficients.
```{r, echo=TRUE}
lambda_vector <- seq(0.1, 1, by = 0.1)
result <- lasso_for_lamda_vector(X_train, y_train, lambda_vector)
beta_matrix <- result$coefficients[-1, ] # exclude intercept
lambda_values <- result$lambda
plot(log(lambda_values), beta_matrix[1, ], type = "l", col = 1, lwd = 2,
     xlab = "log(Lambda)", ylab = "Coefficients", main = "LASSO Coefficient Paths")
for (j in 2:nrow(beta_matrix)) {
  lines(log(lambda_values), beta_matrix[j, ], col = j, lwd = 2)
}

result

mse_test <- numeric(length(lambda_values))
for (i in 1:length(lambda_values)) { # nolint
  lambda <- lambda_values[i]
  result_lasso <- lasso_shooting(X_train, y_train, lambda)
  intercept <- result_lasso$intercept
  beta <- result_lasso$coefficients
  y_pred <- intercept + X_test %*% beta
  mse_test[i] <- mean((y_test - y_pred)^2)
}

# find the best lambda
best_lambda <- lambda_values[which.min(mse_test)]
best_coefficients <- beta_matrix[, which.min(mse_test)]
best_lambda
```

## 2. Compare your fit against the lasso implementation from glmnet.
```{r, echo=TRUE}
result_glmnet <- glmnet(X_train, y_train, alpha = 1, lambda = lambda_vector, standardize = TRUE, intercept = TRUE)
lambda_min_glmnet <- result_glmnet$lambda.min
lambda_min_glmnet
```

## 3. Fit also a ridge regression and a least squares regression for the data (you can use here glmnet).
```{r, echo=TRUE}
cv_fit_ridge <- cv.glmnet(X, y, alpha = 0, nfolds = 10)
lambda_min_ridge <- cv_fit_ridge$lambda.min
best_coef_ridge <- coef(cv_fit_ridge, s = lambda_min_ridge)
lambda_min_ridge

lm_fit <- lm(y ~ ., data = data.frame(X, y))

```

## 4. Compute the lasso, ridge regression and ls regression predictions for the testing data. Which method gives the better predictions? Interpret all three models and argue about their performances.
```{r, echo=TRUE}
# our lasso
lasso_model <- lasso_shooting(X_train, y_train, best_lambda)
lasso_pred <- lasso_model$intercept + X_test %*% lasso_model$coefficients

# ridge
ridge_model <- glmnet(X_train, y_train, alpha = 0, lambda = best_lambda, standardize = TRUE)
ridge_pred <- predict(ridge_model, s = best_lambda, newx = X_test)

# ls
ls_model <- lm(y_train ~ ., data = as.data.frame(X_train))
ls_pred <- predict(ls_model, newdata = as.data.frame(X_test))

# mse for each model
mse_lasso <- mean((y_test - lasso_pred)^2)
mse_ridge <- mean((y_test - ridge_pred)^2)
mse_ls <- mean((y_test - ls_pred)^2)
mse_df <- data.frame(
  Model = c("LASSO", "Ridge Regression", "Least Squares"),
  MSE = c(mse_lasso, mse_ridge, mse_ls)
)
kable(mse_df, caption = "Model comparison MSE")
```

# Task 3

**Regularized** regression means adding a penalty term to the loss function in order to
prevent overfitting. This penalty is supposed to penalize large coefficients meaning 
the model is more likely to generalize easier

**Shrinkage** is a technique that aims to reduce the coefficients of the model to avoid 
overfitting and make the model simpler. To remove the coefficients we introduce a penalty 
term that forces some coefficients to be closer to zero.

**The main difference between Lasso and Ridge** is the penalty term, where lasso uses
the absolute value of beta (L1 regularization), whereas ridge uses beta squared in the penalty term (L2 regularization).
Lasso can drive some coefficients to exactly zero as opposed to ridge which may reduce 
some coefficients closer to zero but not exactly zero, therefore lasso is more focused on 
feature selection (shrinkage). Ridge on the other hand is more focused on using most if not 
all coefficients but in a generalized manner, making sure the model does not have very large coeficients.
