---
title: "Exercise 1 - Introduction to Simulation with Variance Estimation"
author: "Florescu Andrei - 12242987"
date: "2024"
output: pdf_document
---
# Setup

## Set seed

```{r, echo=TRUE}
set.seed(12242987)
```

## Datasets

```{r, echo=TRUE}
x1 <- rnorm(100)
x2 <- rnorm(100, mean = 1000000)
```

# 1. Algorithm function implementations

## Algorithm 1 - (two - pass algorithm - variance calculation in R)
```{r, echo=TRUE}
precise_var <- function(data) {
    sample_mean <- mean(data)
    sum <- sum((data - sample_mean)^2)
    variance <- sum / (length(data) - 1)
    return(variance)
}
```

## Algorithm 2 - (one - pass algorithm - previously variance calculation in Excel)
```{r, echo=TRUE}
excel_var <- function(data) {
    p1 <- sum(data^2)
    p2 <- (sum(data)^2) / length(data)
    variance <- (p1 - p2) / (length(data) - 1)
    return(variance)
}
```

## Algorithm 3 - (shifted one - pass algorithm):
```{r, echo=TRUE}
shifted_var <- function(data, c) {
    p1 <- sum((data - c)^2)
    p2 <- sum(data - c)^2 / length(data)
    variance <- (p1 - p2) / (length(data) - 1)
    return(variance)
}
```


## Algorithm 4 - (online):
```{r, echo=TRUE}
online_var <- function(data) {
    sample_mean <- mean(data[1:2])
    sum <- sum((data[1:2] - sample_mean)^2)
    variance <- sum / (length(data[1:2]) - 1)

    for (i in 3:length(data)) {
        sample_mean1 <- sample_mean
        sample_mean <- sample_mean + (data[i] - sample_mean) / i
        variance <- (i - 2) * variance / (i - 1) + ((data[i] - sample_mean1)^2) / i
    }

    return(variance)
}
```

## Algorithm function wrappper

```{r, echo=TRUE}
compute_variance <- function(data, type) {
    result <- switch(type,
        "precise" = precise_var(data),
        "excel" = excel_var(data),
        "shifted" = shifted_var(data, data[1]),
        "online" = online_var(data),
        "var" = var(data)
    )
    return(result)
}
```

## Call each compute variance

```{r, echo=TRUE}
precise1 <- compute_variance(x1, "precise")
precise2 <- compute_variance(x2, "precise")
excel1 <- compute_variance(x1, "excel")
excel2 <- compute_variance(x2, "excel")
shifted1 <- compute_variance(x1, "shifted")
shifted2 <- compute_variance(x2, "shifted")
online1 <- compute_variance(x1, "online")
online2 <- compute_variance(x2, "online")
rvar1 <- compute_variance(x1, "var")
rvar2 <- compute_variance(x2, "var")
```

### Variance comparison using all.equal

```{r, echo=TRUE}
all.equal(precise1, excel1, shifted1, online1, rvar1)
all.equal(precise2, excel2, shifted2, online2, rvar2)
```

### Variance comparison for x1 using ==

```{r, echo=TRUE}
variables1 <- c(precise1, excel1, shifted1, online1)
names <- c("precise", "excel", "shifted", "online")
comparison_matrix1 <- matrix(variables1 == rvar1, nrow = 1)
colnames(comparison_matrix1) <- names
rownames(comparison_matrix1) <- "R var function"
print(comparison_matrix1)
```

### Variance comparison for x2 using ==

```{r, echo=TRUE}
variables2 <- c(precise2, excel2, shifted2, online2)
comparison_matrix2 <- matrix(variables2 == rvar2, nrow = 1)
colnames(comparison_matrix2) <- names
rownames(comparison_matrix2) <- "R var function"
print(comparison_matrix2)
```

### Variance comparison for x1 using identical

```{r, echo=TRUE}
names <- c("precise", "excel", "shifted", "online")
comparison_matrix1 <- matrix(sapply(variables1, function(x) identical(x, rvar1)), nrow = 1)
colnames(comparison_matrix1) <- names
rownames(comparison_matrix1) <- "R var function"
print(comparison_matrix1)
```

### Variance comparison for x2 using identical

```{r, echo=TRUE}
comparison_matrix2 <- matrix(sapply(variables2, function(x) identical(x, rvar2)), nrow = 1)
colnames(comparison_matrix2) <- names
rownames(comparison_matrix2) <- "R var function"
print(comparison_matrix2)
```

We can see that using == and identical yields the same results for comparing these variances.
We can also see that all.equal returned TRUE for both datasets even though this is not right.
Therefore we can conclude that it's more feasable to use == operator or the identical function
to compare results.
Another thing to note is that the only algorithm that returned the same result as the 
R var function is the shifted algorithm. The precise algorithm returned the same value only once,
and the rest didn't return the same exact result in either case.

```{r, echo=TRUE}
variables1 <- c(variables1, rvar1)
variables2 <- c(variables2, rvar2)
names <- c(names, "R var")
result <- matrix(c(variables1, variables2), nrow = 2)
colnames(result) <- names
rownames(result) <- c("mean=0", "mean=1000000")
print(result)
```

# 2. Compare computational performance

```{r, echo=TRUE, fig.width=6,fig.height=6}
library(microbenchmark)

benchmark_result <- microbenchmark(
    compute_variance(x1, "precise"),
    compute_variance(x1, "excel"),
    compute_variance(x1, "shifted"),
    compute_variance(x1, "online"),
    compute_variance(x1, "var"),
    times = 1000L
)

print(benchmark_result)
par(las = 2)
boxplot(benchmark_result)
```

From the boxplot we can see that the slowest algorithm is the online one while the 
fastest algorithm is the excel one. We can also see that the times are closer to 
each other for precise and online while for the rest they have more scattered runtimes.

# 3. Scale invariance property

```{r, echo=TRUE}
variances <- c(
    shifted_var(x1, x1[1]),
    shifted_var(x1, 1),
    shifted_var(x1, 100000),
    shifted_var(x1, mean(x1)),
    var(x1)
)
comparison_matrix <- matrix(variances, nrow = 1)
names <- c("x1[1]", "1", "1000", "mean", "R var function")
colnames(comparison_matrix) <- names
rownames(comparison_matrix) <- "Variance of x1"
print(comparison_matrix)

benchmark_result <- microbenchmark(
    shifted_var(x1, x1[1]),
    shifted_var(x1, 1),
    shifted_var(x1, 100000),
    shifted_var(x1, mean(x1)),
    var(x1)
)
print(benchmark_result)
par(las = 2)
boxplot(benchmark_result)
```


# 4. Condition numbers

```{r, echo=TRUE}
calculate_condition_number <- function(data) {
    S <- sum((data - mean(data))^2)
    k <- sqrt(1 + (mean(data)^2) * length(data) / S)
    return(k)
}

calculate_condition_number_shifted <- function(data, c) {
    S <- sum((data - mean(data))^2)
    k <- sqrt(1 + ((mean(data) - c)^2) * length(data) / S)
    return(k)
}

calculate_condition_number(x1)
calculate_condition_number(x2)
calculate_condition_number_shifted(x1, mean(x1))
calculate_condition_number_shifted(x2, mean(x1))

x3 <- sample(1:1000000, 100, replace = TRUE)
calculate_condition_number(x3)
calculate_condition_number_shifted(x3, 0.1)
```

</div></pre>